---
title: "Wage Inequality Analysis"
output: word_document
date: "2025-04-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r cars}
filepath1 <- "C:/Users/tineo/OneDrive/Desktop/R/oesm23st/state_M2023_dl.xlsx"
wagesRaw <- readxl::read_excel(filepath1) %>% 
  clean_names()

filepath2 <- "C:/Users/tineo/OneDrive/Desktop/R/oesm23nat/national_M2023_dl.xlsx"
wagesNationalRaw <- readxl::read_excel(filepath2) %>% 
  clean_names()
```

#data cleaning
```{r}
#count how many missing values are there
wagesRaw %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "missing_count") %>%
  arrange(desc(missing_count))


#turn mean and median into numeric from character and remove NAs for these two metrics
wagesCleaned <- wagesRaw %>%
  mutate(a_mean = as.numeric(a_mean), a_median = as.numeric(a_median)) %>%
  mutate(tot_emp = as.numeric(gsub(",", "", tot_emp))) %>%
  filter(!is.na(a_mean), !is.na(a_median), 
    !area_title %in% c('Guam', 'Puerto Rico', 'Virgin Islands'))

#when we need to only look at just state vs state
allOccupations <- wagesCleaned %>%
  filter(occ_title == 'All Occupations')

```


#Map of USA with average wage for each state
```{r}

allOccupations <- allOccupations %>%
  rename(state = area_title)

allOccupations <- allOccupations %>%
  filter(!state == 'District of Columbia')

#US Map that plots the average weight for all states
plot_usmap(data = allOccupations, values = "a_mean", regions = "states") +
  scale_fill_gradient2(
    low = "white", mid = "#8db600", high = "darkgreen", midpoint = 65000,
    name = "Annual Mean Wage (USD)",
    labels = dollar_format()
  ) +
  labs(
    title = "Annual Mean Wage by U.S. State (excluding D.C.)",
    subtitle = "Source: Bureau of Labor Statistics, May 2023"
  ) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 8),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 8),
    plot.caption = element_text(size = 8)
  )


```


#top 5 and bottom 5 states by salary
```{r}
getTopAndBottomStatesData <- function(data, n = 5) {
  rankedStates <- data %>%
    arrange(desc(a_mean))

  topStates <- head(rankedStates$area_title, n)
  bottomStates <- tail(rankedStates$area_title, n)

  selectedStates <- c(topStates, bottomStates)

  # Now filter the full data to keep all rows for these states
  filteredData <- data %>%
    filter(area_title %in% selectedStates)
}

wagesTopAndBottom <- getTopAndBottomStatesData(allOccupations, n = 5)

wagesTopAndBottom <- wagesTopAndBottom %>%
  mutate(group = ifelse(area_title %in% c("Massachusetts", "District of Columbia", 
                                     "New York", "Washington", "New Jersey", "California"),
                        "Top 5", "Bottom 5"))

#boxplot of the distribution of wages per state 
ggplot(wagesTopAndBottom, aes(x = reorder(area_title, a_mean), y = a_mean, fill = group)) +
  geom_col(width = 0.8) +
  geom_text(aes(label = paste0(dollar(a_mean))), 
            hjust = 1.1, color = "white", size = 3) +
  coord_flip() +
  scale_y_continuous(
    labels = dollar, 
    limits = c(0, 110000)
  ) +
  scale_fill_manual(
    values = c("Top 5" = "darkgreen", "Bottom 5" = "#8db600")
  ) +
  labs(
    title = "Top 5 and Bottom 5 States by Annual Mean Wage",
    subtitle = "Source: Bureau of Labor Statistics, May 2023",
    x = "State",
    y = "Annual Mean Wage (USD)",
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.title.position = "plot",
    plot.subtitle = element_text(size = 8),
    axis.title.x = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    legend.position = "none" 
  )

```


#Common or Rare Jobs
```{r}
#most common or rarest jobs across the dataset
getCommonOrRareJobs <- function(dataset, top = TRUE, n = 5) {
  dataset %>%
    filter(o_group == 'detailed', occ_title != "All Occupations") %>%
    group_by(occ_title) %>%
    summarise(totalEmp = sum(tot_emp, na.rm = TRUE)) %>%
    {
      if (top) {
        arrange(., desc(totalEmp)) %>%
          slice(1:n)
      } else {
        arrange(., totalEmp) %>%
          slice(1:n)
      }
    }
}

commonJobs <- getCommonOrRareJobs(wagesCleaned, top = TRUE, n = 5)
rareJobs <- getCommonOrRareJobs(wagesCleaned, top = FALSE, n = 5)

rareJobs <- rareJobs %>%
  mutate(group = "Bottom 5")
commonJobs <- commonJobs %>%
  mutate(group = "Top 5")
commonRareJobs <- bind_rows(commonJobs, rareJobs)


group_labels <- c(
  "Top 5" = "Most Common Jobs",
  "Bottom 5" = "Least Common Jobs"
)

commonJobs$occ_title <- str_wrap(commonJobs$occ_title, width = 25)
rareJobs$occ_title <- str_wrap(rareJobs$occ_title, width = 25)


#Barplot of common and rare jobs
topPlot <- ggplot(commonJobs, aes(x = reorder(occ_title, totalEmp), y = totalEmp)) +
  geom_col(fill = "darkgreen", width = .8) +
  geom_text(aes(label = comma(totalEmp)), 
            hjust = 1.1, color = "white", size = 3) +
  coord_flip() +
  labs(
    title = "Top 5 Most Common Occupations in the U.S.",
    x = "Occupation",
    y = "Total Employment"
  ) +
  scale_y_continuous(labels = scales::comma_format(), limits = c(0, 5000000)) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(hjust = 0, face = "bold", size = 11),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8)
  )

bottomPlot <- ggplot(rareJobs, aes(x = reorder(occ_title, totalEmp), y = totalEmp)) +
  geom_col(fill = "#8db600", width = .8) +
  geom_text(aes(label = comma(totalEmp)), 
            hjust = 1.1, color = "white", size = 3) +
  coord_flip() +
  labs(
    title = "Top 5 Most Rare Occupations in the U.S.",
    x = "Occupation",
    y = "Total Employment"
  ) +
  scale_y_continuous(labels = scales::comma_format(), limits = c(0, 300)) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(hjust = 0, face = "bold", size = 11),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8)
  )

(topPlot) / (bottomPlot) +
  plot_annotation(
    title = "Top and Bottom 5 Jobs by Total Employment",
    subtitle = "Source: Bureau of Labor Statistics, May 2023"
  ) & 
    theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    legend.position = "none" 
  )

```


#THIS IS USING NATIONAL 
```{r}
topPayingJobs <- wagesCleaned %>%
  filter(!occ_title %in% c("All Occupations")) %>%
  filter(!occ_code == '11-1011') %>% #CEO Duplicate
  filter(!occ_code == '29-1151') %>% #nurse duplicate
  arrange(desc(a_mean)) %>%
  slice(1:5)

WorstPayingJobs <- wagesCleaned %>%
  filter(!occ_title %in% c("All Occupations")) %>%
  filter(!occ_code == '35-3023') %>% #Fast food duplicate
  filter(!occ_code == '39-3031') %>% #user
  filter(!occ_code == '35-9031') %>% #Host
  filter(!occ_code == '41-2010') %>% #Cashiers
  arrange(a_mean) %>%
  slice(1:5)

topPayingJobs <- topPayingJobs %>%
  mutate (wageLabel = dollar(a_mean))


WorstPayingJobs <- WorstPayingJobs %>%
  mutate (wageLabel = dollar(a_mean))


#WorstPayingJobs$occ_title <- str_wrap(WorstPayingJobs$occ_title, width = 25)
#topPayingJobs$occ_title <- str_wrap(topPayingJobs$occ_title, width = 25)

#bar blot for paying jobs
topPlot <- ggplot(topPayingJobs, aes(x = reorder(occ_title, a_mean), y = a_mean)) +
  geom_col(fill = "darkgreen", width = .8) +
geom_text(aes(label = wageLabel), 
            hjust = 1.1, color = "white", size = 3.5) +
  coord_flip() +
  labs(
    title = "Best 5 Paid Occupations on Average in the U.S.",
    x = "Occupation",
    y = "Average Wage"
  ) +
  scale_y_continuous(labels = scales::comma_format(), limits = c(0, 400000)) +
  theme_minimal(base_size = 10)


bottomPlot <- ggplot(WorstPayingJobs, aes(x = reorder(occ_title, a_mean), y = a_mean)) +
  geom_col(fill = "#8db600", width = .8) + 
  geom_text(aes(label = wageLabel), 
            hjust = 1.1, color = "white", size = 3.5) +
  coord_flip() +
  labs(
    title = "Worst 5 Paid Occupations on Average in the U.S.",
    x = "Occupation",
    y = "Average Wage"
  ) +
  scale_y_continuous(labels = scales::comma_format(), limits = c(0,100000)) +
  theme_minimal(base_size = 10)

(topPlot) / (bottomPlot) +
  plot_annotation(
    title = "Top and Bottom 5 Jobs by Annual Mean Wage",
    subtitle = "Source: Bureau of Labor Statistics, May 2023"
  ) & 
    theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    legend.position = "none" 
  )


#will be used in the next segment 
topPayingTitles <- topPayingJobs$occ_title %>% unique() %>% as.character()
topPayingTitles <- stringr::str_trim(topPayingTitles)

worstPayingTitles <- WorstPayingJobs$occ_title %>% unique() %>% as.character()
worstPayingTitles <- stringr::str_trim(worstPayingTitles)

```


#See if best/worst paying jobs are proportionally over/under-represented in high/low-wage states
```{r}
calculateJobConcentration <- function(jobTitles, dataset, statesOfInterest, wrapWidth = 30) {
  
  dataset %>%
    filter(o_group == "detailed", occ_title %in% jobTitles) %>%
    {
      selectedJobs <- .
      
      nationalJobTotals <- selectedJobs %>%
        group_by(occ_title) %>%
        summarise(nationalEmp = sum(tot_emp, na.rm = TRUE)) %>%
        ungroup()
      
      StatesJobTotals <- selectedJobs %>%
        filter(area_title %in% statesOfInterest) %>%
        group_by(occ_title) %>%
        summarise(StateEmp = sum(tot_emp, na.rm = TRUE)) %>%
        ungroup()
      
      left_join(StatesJobTotals, nationalJobTotals, by = "occ_title") %>%
        mutate(
          percentInStates = round(100 * StateEmp / nationalEmp, 1),
          occ_title = str_wrap(occ_title, width = wrapWidth)
        )
    }
}


topJobConcentration <- calculateJobConcentration(
  jobTitles = topPayingTitles,
  dataset = wagesCleaned,
  statesOfInterest = c("District of Columbia", "California", "New York")
)

worstJobConcentration <- calculateJobConcentration(
  jobTitles = worstPayingTitles,
  dataset = wagesCleaned,
  statesOfInterest = c("District of Columbia", "California", "New York")
)

worstJobConcentration$occ_title <- str_wrap(worstJobConcentration$occ_title, width = 30)

topPlot <- ggplot(topJobConcentration, aes(x = reorder(occ_title, percentInStates), y = percentInStates)) +
  geom_col(fill = "darkgreen", width = .8) +
  geom_text(aes(label = paste0(round(percentInStates), "%")), 
            hjust = 1.1, color = "white", size = 3) +
  coord_flip() +
  labs(
    title = "High-Paying Jobs",
    x = "Occupation",
    y = "Percent of National Employment"
  ) +
  scale_y_continuous(
    labels = function(x) paste0(round(x), "%"),
    limits = c(0, max(topJobConcentration$percentInStates) + 5)
  ) + theme_minimal(base_size = 10)

bottomPlot <- ggplot(worstJobConcentration, aes(x = reorder(occ_title, percentInStates), 
                                                y = percentInStates)) +
  geom_col(fill = "#8db600", width = .8) +
  geom_text(aes(label = paste0(round(percentInStates), "%")), 
            hjust = 1.1, color = "white", size = 3) +
  coord_flip() +
  labs(
    title = "Low-Paying Jobs",
    x = "Occupation",
    y = "Percent of National Employment"
  ) +
  scale_y_continuous(
    labels = function(x) paste0(round(x), "%"),
    limits = c(0, max(topJobConcentration$percentInStates) + 5)
  ) + theme_minimal(base_size = 10)

(topPlot) / (bottomPlot) +
  plot_annotation(
    title = "Share of U.S. Employment in Top and Bottom 5 Jobs in CA, NY, and DC",
    subtitle = "Source: Bureau of Labor Statistics, May 2023"
  ) & 
    theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 8),
    axis.title.x = element_text(size = 7),
    axis.title.y = element_text(size = 6),
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 5),
    legend.position = "none"  
  )


```


#Difference between Mean and Median by State
```{r}
plotWageGapByState <- function(data, n = 5) {
  # Step 1: Calculate mean-median gap for each state
  meanMedianDiff <- data %>%
    filter(occ_title == "All Occupations") %>%
    mutate(meanMedianGap = a_mean - a_median) %>%
    select(area_title, meanMedianGap)

  # Step 2: Get top and bottom N states
  topStates <- meanMedianDiff %>%
    filter(!is.na(meanMedianGap)) %>%
    slice_max(meanMedianGap, n = n) %>%
    mutate(group = "Top")

  bottomStates <- meanMedianDiff %>%
    filter(!is.na(meanMedianGap)) %>%
    slice_min(meanMedianGap, n = n) %>%
    mutate(group = "Bottom")

  wageGapStates <- bind_rows(topStates, bottomStates)

  # Step 3: Plot
  ggplot(wageGapStates, aes(x = reorder(area_title, meanMedianGap), y = meanMedianGap, 
                            fill = group)) +
    geom_col(width = 0.8) +
    geom_text(aes(label = paste0(dollar(meanMedianGap))), 
            hjust = 1.1, color = "white", size = 3) +
    coord_flip() +
    scale_fill_manual(
      values = c("Top" = "darkgreen", "Bottom" = "#8db600"),
      guide = "none"  # remove legend if you don't want it
    ) +
    scale_y_continuous(
      labels = scales::dollar_format(),
      limits = c(0, 25000)
    ) +
    labs(
      title = "States with the Largest and Smallest Gaps Between Mean and Median Wages",
      subtitle = "Source: Bureau of Labor Statistics, May 2023",
      x = "State",
      y = "Mean - Median Wage ($)"
    ) +
    theme_minimal(base_size = 11) +
    theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.title.position = "plot",
    plot.subtitle = element_text(size = 8),
    axis.title.x = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    legend.position = "none" 
    )
}


plotWageGapByState(wagesCleaned, n = 5)
```



#What is the occupation that is seen most across these states? 
```{r}

getTopOccupationsInStates <- function(dataset, statesOfInterest, topN = 5, wrapWidth = 30) {
  
  dataset %>%
    filter(area_title %in% statesOfInterest, o_group == "major") %>%
    group_by(occ_title) %>%
    summarise(totalEmp = sum(tot_emp, na.rm = TRUE)) %>%
    mutate(
      percentOfStateEmployment = 100 * totalEmp / sum(totalEmp),
      occ_title = str_wrap(occ_title, width = wrapWidth)
    ) %>%
    arrange(desc(percentOfStateEmployment)) %>%
    slice(1:topN)
}

topJobsLowIncomeStates <- getTopOccupationsInStates(wagesCleaned,
                                                    c("Arkansas", "West Virginia",  "Mississippi"))

topJobsHighIncomeStates <- getTopOccupationsInStates(wagesCleaned,
                                                    c("District of Columbia", "New York",  "California"))


topPlot <- ggplot(topJobsHighIncomeStates, aes(x = reorder(occ_title, percentOfStateEmployment), y = percentOfStateEmployment)) +
  geom_col(fill = "darkgreen", width = .8) +
  geom_text(aes(label = paste0(round(percentOfStateEmployment), "%")), 
            hjust = 1.1, color = "white", size = 3) +
  coord_flip() +
  labs(
    title = "Occupation % in CA, NY, and DC",
    x = "Occupation",
    y = "Percent of State Employment"
  ) +
  scale_y_continuous(
    labels = function(x) paste0(round(x), "%"),
    limits = c(0, max(topJobsLowIncomeStates$percentOfStateEmployment) + 5)
  ) +
  theme_minimal(base_size = 11) 

bottomPlot <- ggplot(topJobsLowIncomeStates, aes(x = reorder(occ_title, percentOfStateEmployment), y = percentOfStateEmployment)) +
  geom_col(fill = "#8db600", width = .8) +
  geom_text(aes(label = paste0(round(percentOfStateEmployment), "%")), 
            hjust = 1.1, color = "white", size = 3) +
  coord_flip() +
  labs(
    title = "Occupation % in AR, WV, and MS",
    x = "Occupation",
    y = "Percent of State Employment"
  ) +
  scale_y_continuous(
    labels = function(x) paste0(round(x), "%"),
    limits = c(0, max(topJobsLowIncomeStates$percentOfStateEmployment) + 5)
  ) +
  theme_minimal(base_size = 11) 

(topPlot) / (bottomPlot) +
  plot_annotation(
    title = "Top Occupations by Employment Share in High and Low Income States",
    subtitle = "Source: Bureau of Labor Statistics, May 2023"
  ) & 
    theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 8),
    axis.title.x = element_text(size = 7),
    axis.title.y = element_text(size = 6),
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 5),
    legend.position = "none" 
  )


```


#Linear Regression - does an increase in average wage increase the wage inequality?
```{r}

#we are only interested in the annual wage per state so we will use the aggregate of all 
#occupations. We will create a new column with mean - median as our dependent variable
wageGapData <- wagesCleaned %>%
  filter(occ_title == "All Occupations") %>%
  mutate(
    wageGap = a_mean - a_median
  ) %>%
  select(area_title, a_mean, a_median, wageGap, tot_emp)


#initial linear regression 
wageGapModel <- lm(wageGap ~ a_mean, data = wageGapData, weights = tot_emp)
summary(wageGapModel)

#Check for Regression assumptions
par(mfrow = c(2,2))
plot(wageGapModel2)

#removing points that have high leverage in the models - validate findings
wageGapDataClean <- wageGapData[-c(5, 9, 33), ]

wageGapModel2 <- lm(formula = wageGap ~ a_mean, data = wageGapDataClean, weights = tot_emp)
summary(wageGapModel2)

```
